Beyond Ultra-Diffuse Galaxies I: Mass-Size Outliers Among the Satellites of Milky Way Analogs

### 2022-10-15 paper1.tex
cannot compile

\author[0000-0003-4970-2874]{Johnny Greco}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}
\affiliation{Center for Cosmology and AstroParticle Physics (CCAPP), The Ohio State University, Columbus, OH 43210, USA}

@@

\author[0000-0002-1841-2252]{Rachael Beaton}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}
\author[0000-0002-2991-9251]{Kirsten Casey}
\affiliation{Center for Cosmology and AstroParticle Physics (CCAPP), The Ohio State University, Columbus, OH 43210, USA}
\author[0000-0002-1841-2252]{Shany Danieli}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}
\author[0000-0002-1841-2252]{Andy Goulding}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}
\author[0000-0002-2704-5028]{Remy Joseph}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}
\affiliation{Oskar Klein Centre for Cosmoparticle Physics, Department of Physics, Stockholm University, Stockholm SE-106 91, Sweden}
\author[0000-0002-1841-2252]{Erin Kado-Fong}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}

@@

\section{Mass-size outliers in Milky-Way analogs}\label{sec:sample_construction}
The goal of this paper is to study the mass-size outliers in the satellite systems around MW-like galaxies. However, distance information is needed to convert the observed size and magnitude to the physical size and stellar mass. It is well-known that getting the distances to LSBGs is one of the major obstacles for studying their properties. Beside direct distance measurements (e.g., SAGA, ELVES, \citealt{Kadowaki2021}), it is common to assume a distance to a LSBG by associating it with a host galaxy based on the projected angular distance \citep[e.g.,][]{vanDokkum2015,vdBurg2016,Wang2021,Zaritsky2022,Nashimoto2022}. A statistical background subtraction is then needed to account for the contribution from background and foreground galaxies. Furthermore, cross-correlation between an LSBG sample and a host galaxy sample can also reveal the distance distribution of LSBGs \citep{Greene2022}. These methods are also complemented by recent machine learning techniques \citep{Baxter2021,xSAGA-I}. In this Section, we cross-match our LSBG catalog with MW-like galaxies in the NSA catalog. For the LSBGs matched with MW analogs, we run the deblending step to exclude false positives (\S\ref{sec:deblending}) and then run Spergel modeling to measure the properties of LSBGs (\S\ref{sec:modeling}). In the end, we construct the samples of mass-size outliers in the satellites of MW analogs.

\subsection{Matching with Milky-Way analogs}\label{sec:match}
The properties of the Milky Way itself vary in the literature \citep{Licquia2015,Bland-Hawthorn2016} and the definitions of MW analogs are different among different studies. In the SAGA survey \citep{SAGA-I,SAGA-II}, MW analogs are selected based on their absolute $K$-band magnitude $-23 > M_K > -24.6$~mag, which is derived using abundance matching by assuming a simple galaxy-halo connection model (see Fig. 2 in \citealt{SAGA-I}). This luminosity range approximately corresponds to a stellar mass range of $10.2 < \log\, M_\star/M_\odot < 11.0$. SAGA also requires that the MW analogs be in isolation (without nearby bright galaxies) and lie in a redshift range of $0.005 < z < 0.01$ (20--40 Mpc). In the ELVES survey \citep{ELVES-I,ELVES-II,CarlstenELVES2022}, the requirements for MW-like host is loosened to be $M_K < -22.1$~mag ($M_\star > 10^{9.9}\ M_\odot$) because the probed volume by ELVES ($D<12$ Mpc) is smaller than that of SAGA. In this work, we choose the stellar mass range of MW analogs to be $10.2 < \log\, M_\star/M_\odot < 11.2$, which is simply a 1 dex bin centered at the measured stellar mass of the Milky Way ($M_{\star, \mathrm{MW}}\approx 10^{10.7}\ M_\odot$, \citealt{Licquia2015}). MW analogs selected using this definition are very close to those in SAGA but are slightly less massive than the ELVES hosts since ELVES has several groups more massive than SAGA hosts.

@@

However, there are still a significant fraction of spurious objects in this cross-matched LSBG sample, including galactic cirrus, tidal streams, shredded large galaxies, and compact sources blended with galaxy outskirts. As described in \S\ref{sec:deblending}, we perform a deblending step to effectively remove these spurious objects. There are 2,673 objects left after the deblending cuts. Next, we model them using the Spergel profile as described in \S\ref{sec:modeling} and obtain their photometry and structural parameters from the best-fit models. 
% Based on the modeling results, we apply the same color cuts as in Section \ref{sec:deblending} and a size cut of $1.6\arcsec < r_e < 15\arcsec$ to further remove unreliable fits and false positives. 
We also remove duplicated objects. In the end, we have 2,510 LSBG candidates as our final LSBG sample. We note that these LSBGs are matched to MW hosts in projection but do not have direct distance measurements.

@@

\subsection{Mass-size outlier sample}\label{sec:sample}
The concept of UDG was proposed to describe dwarf galaxies with unusually large size ($r_e>1.5$ kpc) and diffuse light distribution ($\sbcen > 24.0\ \sbunit$). Although UDGs are large in size, the constant size cut does not account for the fact that the galaxy size is strongly correlated with its mass, known as the mass-size relation \citep[e.g.,][]{Graham2003,Trujillo2007,vanDokkum2013,Cappellari2013,Lange2015}.

A more intuitive definition for diffuse and puffy dwarf galaxies (i.e., mass-size outliers) should be based on a mass-size relation and its scatter. Following this thread, we propose a new definition for mass-size outliers dubbed as \textit{ultra-puffy galaxies} (UPGs). In this section, we describe the definitions of UPG and contrast it with UDG, then we select UDGs and UPGs from our LSBGs matched with MW hosts. 
% Ideally, galaxies that are $1.5\sigma$ or $2\sigma$ above the average mass-size relation represent the large-size tail of the population in a homogeneous manner.

\subsubsection{UPGs}
To select mass-size outliers, a mass-size relation and its scatter are needed. However in general, the slope of the mass-size relation in the nearby Universe has been shown to be color- and morphology-dependent for galaxies above $M_\star > 10^{9}\ M_\odot$: blue star-forming galaxies have a shallower slope than the red quenched galaxies \citep[e.g.,][]{Lange2015}. Fortunately, in the dwarf galaxy regime ($10^{5.5} < M_\star < 10^{8.5}$), \citet{ELVES-I} derive the mass-size relation from the satellites of MW analogs in the Local Volume and show that the average mass-size relation is quite universal: the slope and intercept are not sensitive to the color or morphology of dwarf galaxies. The residual from the mass-size relation follows a log-normal distribution reasonably well. This gives us a chance to define a subset of dwarf galaxies that are size-outliers with respect to the average mass-size relation. It is worth noting that a definition based on the mass-size relation was already used in simulation studies where simulations cannot reproduce the observed mass-size relation \citep[e.g.,][]{Benavides2021,Benavides2022}.

Taking the measured average mass-size relation from \citet{ELVES-I} $\log\, (r_e/\mathrm{pc}) = 0.25\, \log\, (M_\star/M_\odot) + 1.07$ and a scatter of $\sigma=0.18$ dex, we define a population of ultra-puffy galaxies to be galaxies that are $>1.5\sigma$ above the average mass-size relation. We choose $1.5\sigma$ just because it gives us a statistically significant sample. One can certainly select $>2\sigma$ or $>3\sigma$ UPGs if a larger LSBG sample is available. The UPGs are not absolutely large in size or absolutely faint in surface brightness, but they are outliers in size for their stellar masses. The definition of UPG based on the observed mass-size relation is more physically-motivated, and better allows us to study size outliers at a given stellar mass. We further discuss the advantages and disadvantages of UDG and UPG in \S\ref{sec:mass-size} and \S\ref{sec:discussion}.

@@

We randomly select a continuous patch of sky of 24 deg$^{2}$ in HSC PDR2 regardless of whether containing MW analogs\footnote{$345\ \deg < \rm{R.A.} < 351\ \deg$, $-1.5\ \deg < \rm{Dec} < 2.5\ \deg$, which is not dominated by Galactic cirrus and stars.}. Then we perform the same deblending and modeling steps for the 2,707 LSBGs detected in this region. We also removed objects that are already in our mass-size outlier samples and objects with completeness less than 0.1. Furthermore, we also did the same visual inspection as for UDGs and UPGs to remove false positives. In the end, there are 480 LSBGs (excluding false positives) representing a population of possible contaminants for the UDG and UPG samples. Next, because both UDG and UPG are defined based on a physical size, we randomly associate the 480 LSBGs to the hosts of our UDGs and UPGs, respectively. Then we calculate a physical size for each LSBG and classify an LSBG as a ``fake`` UDG or UPG if it satisfies the corresponding definition. We repeat such random matching 200 times. In the end, we obtain 11,458 fake UDGs and 11,867 fake UPGs. These UDGs and UPGs are ``fake'' only in the sense of being associated with random MW hosts.

@@

\section{Results}\label{sec:results}
In this section, we present a statistical analysis for mass-size outliers in the satellites of MW analogs at $0.01 < z < 0.04$. We first compare the distributions of UDGs and UPGs on the size-surface brightness plane and mass-size plane to gain intuition on the differences between UDG and UPG definitions. Then we present the abundances of UDGs and UPGs and compare with literature.

@@

In order to build a more comprehensive understanding on our UDG and UPG sample, we show their distributions on the size-surface brightness plane in Figure \ref{fig:udg_upg_re_mu}. The galaxies are split into two color bins and are shown in blue ($g-i < 0.8$) and red ($g-i > 0.8$), respectively. We note that we do not apply any background contamination correction in Figure \ref{fig:udg_upg_re_mu} because such correction can only be done in a statistical sense. The error bars correspond to $1\sigma$ measurement uncertainties (\S\ref{sec:comp_meas}). Some galaxies with size smaller than 1.5 kpc are still scattered in the UDG sample because their sizes have large uncertainty. The two marginal plots show the unnormalized histograms of galaxies in the two color bins. The numbers of red and blue galaxies are similar in the UDG sample, but there are more blue galaxies than red ones in the UPG sample. Since there is no hard surface brightness cut for UPGs, the UPG sample includes many blue galaxies with brighter surface brightness ($23 \lesssim \sbeff < 25\ \sbunit$). Due to the surface brightness cut ($\sbeff > 23\ \sbunit$) in the deblending step, there is little chance for a LSB disk galaxy with bright bulge to scatter in the UPG sample. The UPG sample also loses a significant fraction of red galaxies at $25 < \sbeff < 26\ \sbunit$ because their stellar masses are too high for their sizes to make them being size-outliers.

@@

\jiaxuan{Discuss whether UPGs are LSB disks?}

% discriminative for blue and red galaxies.

@@

We define the UDG (UPG) abundance as the average number of UDGs (UPGs) per host galaxy. In our UDG sample, there are 412 UDGs associated with 258 hosts. After correcting for background contamination (see \S \ref{sec:bkg}) and completeness (see \S \ref{sec:comp_meas}), the UDG abundance of MW analogs is $N_{\rm UDG} = 0.76\pm 0.04$ per host. The UPG abundance in our sample is $N_{\rm UPG} = 0.61\pm 0.04$ per host. Here we neglect the fact that the number of satellites contained within the virial sphere is different from the number of satellite within a projected cylinder with virial radius (the so-called deprojection factor, \citealt{vdBurg2017}).

@@

>>>>

\author[0000-0003-4970-2874]{Johnny P. Greco}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}
\affiliation{Center for Cosmology and AstroParticle Physics (CCAPP), The Ohio State University, Columbus, OH 43210, USA}

@@

\author[0000-0002-1841-2252]{Rachael Beaton}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}
\author[0000-0002-2991-9]{Kirsten Casey}
\affiliation{Center for Cosmology and AstroParticle Physics (CCAPP), The Ohio State University, Columbus, OH 43210, USA}
\author[0000-0002-1841-2252]{Shany Danieli}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}
\author[0000-0002-1841-2252]{Andy Goulding}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}
\author[0000-0002-2704-5028]{Remy Joseph}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}
\affiliation{Oskar Klein Centre for Cosmoparticle Physics, Department of Physics, Stockholm University, Stockholm SE-106 91, Sweden}
\author[0000-0002-1841-2252]{Erin Kado-Fong}
\affiliation{Department of Astrophysical Sciences, 4 Ivy Lane, Princeton University, Princeton, NJ 08544, USA}

@@

\section{Mass-size outliers in Milky-Way analogs}\label{sec:sample_construction}
The goal of this paper is to study the mass-size outliers in the satellite systems around MW-like galaxies. However, distance information is needed to convert the observed size and magnitude to the physical size and stellar mass. It is well-known that getting the distances to LSBGs is one of the major obstacles for studying their properties. Beside direct distance measurements (e.g., SAGA, ELVES, \citealt{Kadowaki2021}), it is common to assume a distance to a LSBG by associating it with a host galaxy based on the projected angular distance \citep[e.g.,][]{vanDokkum2015,vdBurg2016,Wang2021,Zaritsky2022,Nashimoto2022}. A statistical background subtraction is then needed to account for the contribution from background and foreground galaxies. Furthermore, cross-correlation between an LSBG sample and a host galaxy sample can also reveal the distance distribution of LSBGs \citep{Greene2022}. These methods are also complemented by recent machine learning techniques \citep{Baxter2021,xSAGA-I}. In this Section, we cross-match our LSBG catalog with MW-like galaxies in the NSA catalog. For the LSBGs matched with MW analogs, we run the deblending step to exclude false positives (\S\ref{sec:deblending}) and then run Spergel modeling to measure the properties of LSBGs (\S\ref{sec:modeling}). In the end, we construct the samples of mass-size outliers among the satellites of MW analogs.

\subsection{Matching with Milky-Way analogs}\label{sec:match}
The properties of the Milky Way itself vary in the literature \citep{Licquia2015,Bland-Hawthorn2016} and the definitions of MW analogs are different among different studies. In the SAGA survey \citep{SAGA-I,SAGA-II}, MW analogs are selected based on their absolute $K$-band magnitude $-23 > M_K > -24.6$~mag (corresponding to a stellar mass range of $10.2 < \log\, M_\star/M_\odot < 11.0$) and redshift $0.005 < z < 0.01$ (20--40 Mpc). SAGA also requires that the MW analogs be in isolation (without nearby bright galaxies). In the ELVES survey \citep{ELVES-I,ELVES-II,CarlstenELVES2022}, the requirements for MW-like host is loosened to be $M_K < -22.1$~mag ($M_\star > 10^{9.9}\ M_\odot$) because the probed volume by ELVES ($D<12$ Mpc) is smaller than that of SAGA. In this work, we choose the stellar mass range of MW analogs to be $10.2 < \log\, M_\star/M_\odot < 11.2$, which is simply a 1 dex bin centered at the measured stellar mass of the Milky Way ($M_{\star, \mathrm{MW}}\approx 10^{10.7}\ M_\odot$, \citealt{Licquia2015}). MW analogs selected using this definition are very close to those in SAGA but are slightly less massive than the ELVES hosts since ELVES has several groups more massive than SAGA hosts.

@@

However, there are still a significant fraction of spurious objects in this cross-matched LSBG sample, including galactic cirrus, tidal streams, shredded large galaxies, and compact sources blended with galaxy outskirts. As described in \S\ref{sec:deblending}, we perform a deblending step to effectively remove these spurious objects. There are 2,673 objects left after the deblending cuts. Next, we model them using the Spergel profile as described in \S\ref{sec:modeling} and obtain their photometry and structural parameters from the best-fit models. 
% Based on the modeling results, we apply the same color cuts as in Section \ref{sec:deblending} and a size cut of $1.6\arcsec < r_e < 15\arcsec$ to further remove unreliable fits and false positives. 
We also remove duplicated objects. In the end, we have 2,510 LSBG candidates associated with 689 MW analogs as our final LSBG sample. We note that these LSBGs are matched to MW hosts in projection but do not have direct distance measurements.

@@

\subsection{Mass-size outlier sample}\label{sec:sample}
The concept of UDG was proposed to describe dwarf galaxies with unusually large size ($r_e>1.5$ kpc) and diffuse light distribution ($\sbcen > 24.0\ \sbunit$). Although UDGs are large in size, the constant size cut does not account for the fact that the galaxy size is strongly correlated with its mass, known as the mass-size relation \citep[e.g.,][]{Graham2003,Trujillo2007,vanDokkum2013,Cappellari2013,Lange2015}. For example, a galaxy with $r_e = 1.5$ kpc is very extended if its stellar mass is $10^7\ M_\odot$, but is normal-sized if its stellar mass is $10^{8.5}\ M_\odot$.
Thus, a more intuitive definition for diffuse and puffy dwarf galaxies (i.e., mass-size outliers) should be based on a mass-size relation and its scatter. Following this thread, we propose a new definition for mass-size outliers dubbed as \textit{ultra-puffy galaxies} (UPGs). In this section, we describe the definitions of UPG and contrast it with UDG, then we select UDGs and UPGs from our LSBGs matched with MW hosts. 
% Ideally, galaxies that are $1.5\sigma$ or $2\sigma$ above the average mass-size relation represent the large-size tail of the population in a homogeneous manner.

\subsubsection{UPGs}
To select mass-size outliers, a mass-size relation and its scatter are needed. However in general, the slope of the mass-size relation in the nearby Universe has been shown to be color- and morphology-dependent for galaxies above $M_\star > 10^{9}\ M_\odot$: blue star-forming galaxies have a shallower slope than the red quenched galaxies \citep[e.g.,][]{Lange2015}. Fortunately, in the dwarf galaxy regime ($10^{5.5} < M_\star < 10^{8.5}$), \citet{ELVES-I} derive the mass-size relation from the satellites of MW analogs in the Local Volume and show that the average mass-size relation is quite universal: the slope and intercept are not sensitive to the color or morphology of dwarf galaxies. The residual from the mass-size relation follows a log-normal distribution reasonably well. This gives us a chance to define a subset of dwarf galaxies that are size-outliers with respect to the average mass-size relation. It is worth noting that a definition based on the mass-size relation was already used in simulation studies where the absolute sizes were uniformly too large such that size outliers can only be selected relative to the simulated mass-size relation \citep[e.g.,][]{Benavides2021,Benavides2022}.

Taking the measured average mass-size relation from \citet{ELVES-I} $\log\, (r_e/\mathrm{pc}) = 0.25\, \log\, (M_\star/M_\odot) + 1.07$ and a scatter of $\sigma=0.18$ dex, we define a population of ultra-puffy galaxies to be galaxies that are $>1.5\sigma$ above the average mass-size relation. We choose $1.5\sigma$ just because it gives us a statistically significant sample. One can certainly select $>2\sigma$ or $>3\sigma$ UPGs if a larger LSBG sample is available. The UPGs are not absolutely large in size or absolutely faint in surface brightness, but they are outliers in size for their stellar masses\footnote{A UPG with $M_\star = 10^7\ M_\odot$ has a size larger than 1.2 kpc and surface brightness fainter than $\sbeff > 25.5\ \sbunit$ for $g-i=0.4$ and $\sbeff > 26.8\ \sbunit$ for $g-i=0.8$. A UPG with $M_\star = 10^{8.5}\ M_\odot$ has a size larger than 2.7 kpc and surface brightness fainter than $\sbeff > 23.7\ \sbunit$ for $g-i=0.4$ and $\sbeff > 25.0\ \sbunit$ for $g-i=0.8$.}. The definition of UPG based on the observed mass-size relation is more physically-motivated, and better allows us to study size outliers at a given stellar mass. We further discuss the advantages and disadvantages of UDG and UPG in \S\ref{sec:mass-size} and \S\ref{sec:discussion}.

@@

We randomly select a continuous patch of sky of 24 deg$^{2}$ in HSC PDR2 regardless of whether containing MW analogs\footnote{$345\ \deg < \rm{R.A.} < 351\ \deg$, $-1.5\ \deg < \rm{Dec} < 2.5\ \deg$, which is not dominated by Galactic cirrus and stars. Such a patch of sky with 24 deg$^2$ is already enough for us to derive a background contamination fraction with small Poisson error.}. Then we perform the same deblending and modeling steps for the 2,707 LSBGs detected in this region. We also removed objects that are already in our mass-size outlier samples and objects with completeness less than 0.1. Furthermore, we also did the same visual inspection as for UDGs and UPGs to remove false positives. In the end, there are 480 LSBGs (excluding false positives) representing a population of possible contaminants for the UDG and UPG samples. Next, because both UDG and UPG are defined based on a physical size, we randomly associate the 480 LSBGs to the hosts of our UDGs and UPGs, respectively. Then we calculate a physical size for each LSBG and classify an LSBG as a ``fake`` UDG or UPG if it satisfies the corresponding definition. We repeat such random matching 200 times. In the end, we obtain 11,458 fake UDGs and 11,867 fake UPGs. These UDGs and UPGs are ``fake'' only in the sense of being associated with random MW hosts.

@@

\section{Results}\label{sec:results}
In this section, we present a statistical analysis for mass-size outliers among the satellites of MW analogs at $0.01 < z < 0.04$. We first compare the distributions of UDGs and UPGs on the size-surface brightness plane and mass-size plane to gain intuition on the differences between UDG and UPG definitions. Then we present the abundances of UDGs and UPGs and compare with literature.

@@

In order to build a more comprehensive understanding on our UDG and UPG sample, we show their distributions on the size-surface brightness plane in Figure \ref{fig:udg_upg_re_mu}. The galaxies are split into two color bins and are shown in blue ($g-i < 0.8$) and red ($g-i > 0.8$), respectively. We note that we do not apply any background contamination correction in Figure \ref{fig:udg_upg_re_mu} because such correction can only be done in a statistical sense. The error bars correspond to $1\sigma$ measurement uncertainties (\S\ref{sec:comp_meas}). Some galaxies with size smaller than 1.5 kpc are still scattered in the UDG sample because their sizes have large uncertainty. The two marginal plots show the unnormalized histograms of galaxies in the two color bins. The numbers of red and blue galaxies are similar in the UDG sample, but there are more blue galaxies than red ones in the UPG sample. Since there is no hard surface brightness cut for UPGs, the UPG sample includes many blue galaxies with brighter surface brightness ($23 \lesssim \sbeff < 25\ \sbunit$). For blue UPGs (shown as blue dots), the correlation between $\sbeff$ and $r_e$ is merely the fact that blue UPGs pile up around the $1.5\sigma$ line above the mass-size relation.

Due to the surface brightness cut ($\sbeff > 23.0\ \sbunit$) in the deblending step, there is little chance for a LSB disk galaxy with bright bulge to scatter in the UPG sample. The UPG sample also loses a significant fraction of red galaxies at $25 < \sbeff < 26\ \sbunit$ because their stellar masses are high but their sizes are too small to be mass-size outliers.

@@

@@

We define the UDG (UPG) abundance as the average number of UDGs (UPGs) per host galaxy. We searched 901 MW analogs. In our UDG sample, there are 412 UDGs associated with 258 hosts. After correcting for background contamination (see \S \ref{sec:bkg}) and completeness (see \S \ref{sec:comp_meas}), the UDG abundance of MW analogs is $N_{\rm UDG} = 0.76\pm 0.04$ per host. The UPG abundance in our sample is $N_{\rm UPG} = 0.61\pm 0.04$ per host. Here we neglect the fact that the number of satellites contained within the virial sphere is different from the number of satellite within a projected cylinder with virial radius (the so-called deprojection factor, \citealt{vdBurg2017}).

@@

We note that ELVES probes less massive dwarfs than our search.
